Assimilating disturbance: Toward real-time monitoring and forecasting

### 2023-02-01 methods.Rmd
Discussion: Simulated data and OR test case text
formatting

```{r concept, echo=FALSE, message = FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
library(rjags)
par(mfrow=c(2,1))
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

For the second experiment, we used the VSEM to extend the Multinomial assimilation model to a single site, single disturbance example with multiple pools. In this and the following simulated data experiments we used VSEM's default parameters, drawing ensemble members from Normal distributions centered on the prior mean and assuming a standard deviation that was 1/10th the default uniform prior range. The goal here was not to replicate any specific system, but to provide an amount of parameter uncertainty that was realistic but not so large as to generate trivial ensemble forecasts. Model runs were for 25 years using VSEM's default simulated PAR as the driver and with data assimilation occurring annually. Model initial conditions were generated by an initial 25 year ensemble spin-up prior to the 25 year experiment. Process error was added to the simulation annually assuming a zero-bound tobit distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools. Ensemble members experienced disturbance stochastically with a 5% annual probability. For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 [UNITS?] for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).

@@

Our study region was a Landsat tile over central Oregon, USA located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude. For this tile we extracted data from three derived Landtrendr datasets (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html, tile h03v04):  biomass, cover, and disturbance. All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover. The biomass product calibrated Landtrendr-processed Landsat data against field plot data using the gradient nearest neighbor (GNN) methodology to produce an annual biomass estimates [@Kennedy2018]. The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S. National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018]. To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_). The disturbance product **need text and citation from Robert** [CITE]. To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.

```{r map, echo=FALSE, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
## maps of pre-disturbance biomass, cover, and 2004 disturbance

@@

## Biomass
if(!exists("bio04")) bio04 = raster::raster("bio04.tif")
raster::plot(bio04,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main="Landtrendr Biomass")

@@

For the assimilation experiment, each site was assimilated separately. In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^2$ based on [@Kennedy2018] Figure 2. Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017). The difference between the predicted and observed biomass at the end of the assimilation was taken as a metric of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations. **ONCE RESULTS ARE FINALIZED, COME BACK AND FILL IN WHICH SPECIFIC ANALYSES WERE USED**.

Finally, based on preliminary results we performed a series of additional assimilation experiments where we manually altered the observation error, R, or process error, Q, to assess the sensitivity of our results. Table (\@ref(tab:QRexp)) lists the multipliers that were used to adjust R and Q. The skill of these assimilation experiments were assessed the same as the original. In addition, changes in skill across experiments were assessed as a function of R and Q.

>>>>

```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
library(rjags)
par(mfrow=c(2,1))
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

For the second experiment, we used the VSEM to extend the Multinomial assimilation model to a single site, single disturbance example with multiple pools. In this and the following simulated data experiments we used VSEM's default parameters, drawing ensemble members from Normal distributions centered on the prior mean and assuming a standard deviation that was 1/10th the default uniform prior range. The goal here was not to replicate any specific system, but to provide an amount of parameter uncertainty that was realistic but not so large as to generate trivial ensemble forecasts. Model runs were for 25 years using VSEM's default simulated PAR as the driver and with data assimilation occurring annually. Model initial conditions were generated by an initial 25 year ensemble spin-up prior to the 25 year experiment. Process error was added to the simulation annually assuming a zero-bound tobit distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools. Ensemble members experienced disturbance stochastically with a 5% annual probability. For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 $kgC/m^2$ for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).

@@

Our study region was a Landsat tile over central Oregon, USA located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude. For this tile we extracted data from three derived Landtrendr datasets (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html, tile h03v04):  biomass, cover, and disturbance. All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover. The biomass product calibrated Landtrendr-processed Landsat data against field plot data using the gradient nearest neighbor (GNN) methodology to produce an annual biomass estimates [@Kennedy2018]. The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S. National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018]. To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_). 
The disturbance product **need text from Robert** [@KENNEDY2010]. 
To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.

```{r map, echo=FALSE, fig.asp=1, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
## maps of pre-disturbance biomass, cover, and 2004 disturbance

@@

## Biomass
if(!exists("bio04")) bio04 = raster::raster("bio04.tif")
raster::plot(bio04/10,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main=expression("Landtrendr Biomass" ~ (kgC/m^2)))

@@

For the assimilation experiment, each site was assimilated separately. In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^2$ based on [@Kennedy2018] Figure 2. Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017). The difference between the predicted and observed biomass at the end of the assimilation was taken as a metric of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations. To standardize this into a unitless metric, the difference in biomass was divided by half the width of the Analysis 95% CI. Across sites the extent of filter divergence was analyzed as a function of disturbance type and post-disturbance AGB using linear models. Similarly, the ability of the algorithm to detect disturbance, represented as the mean posterior estimate of the disturbance probability, was assessed as a function of disturbance type and post-disturbance AGB using generalized linear models with a logit link and quasibinomial family.

Finally, based on preliminary results we performed a series of additional assimilation experiments where we manually altered the observation error, R, or process error, Q, to assess the sensitivity of our results. Table (\@ref(tab:QRexp)) lists the multipliers that were used to adjust R and Q. The skill of these assimilation experiments were assessed the same as the original. In addition, changes in skill and filter divergence across experiments were assessed as a function of R and Q focusing on two response variables: (1) the filter coverage, defined as the proportion of sites where our divergence metric falls between -1 and 1 (i.e., where the difference between the observation and Analysis is less than half the width of the Analysis 95% CI); and (2) the detection threshold, defined as the change in AGB where the probability of detecting a disturbance is 50% as inferred by the GLM of disturbance size versus probability of disturbance.
